Sex-Specific Evolution of the Meiotic Recombination Rate

### 2020-07-01 doc/MS_v4.Rmd
citation for physical linkage

Meiosis converts diploid germ cells into haploid gametes. During meiosis I, DNA crossovers aid the separation of homologous chromosomes by physically linking them and establishing tension between them on the spindle **(REF?)**. The wrong number of recombination events can disrupt chromosomal segregation, leading to infertility, miscarriage, and birth defects [@hassold2001]. Recombination also shapes evolution by shuffling the combinations of genetic variants offspring inherit. Recombination affects the fates of beneficial and deleterious mutations [@felsenstein1974; @fisher1930; @hill1966] and interacts with natural selection to leave gradients in genomic patterns of diversity [@begun1992; @nachman2012; @charlesworth1993; @cutter2013; @smith1974].

@@

We used counts of MLH1 foci per cell to estimate genome-wide recombination rates in `r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` -- 14 wild-derived inbred strains sampled from three subspecies of house mice ( _M. m. domesticus_, _M. m. musculus_ and _M. m. molossinus_ ) and three other species of Mus ( _M. spretus_, _M. spicilegus_ , and _M. caroli_ ). Mean MLH1 focus counts for `r sum(main.table$Nmice)` -- 188 mice were quantified from an average of `r mean( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T)` -- 21.77 spermatocytes per male (for a total of `r formatC( sum( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T), format="d", big.mark=",")` -- 1,742 spermatocytes) and `r mean( (Mouse.Table_HQ %>% filter(sex == "female") )$Ncells, na.rm = T)` -- 17.85 oocytes per female (for a total of `r formatC(sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells), format="d", big.mark=",")` -- 1,427 oocytes). **<insert statement for cell sample size number>**

Graphical comparisons reveal sex-specific dynamics to the evolution of genome-wide recombination rate (Figure 1A). First, MLH1 focus counts differ between females and males in most strains. Second, the difference in counts between the sexes varies among strains. Although most strains show more MLH1 foci in females, two strains (*musculus^PWD^* and _molossinus^MSM^_) exhibit higher counts in males. In females, numbers of MLH1 foci are evenly distributed around the sex-wide mean of approximately 25 (Figure 1B). In stark contrast, males largely separate into two groups of strains with high numbers (near 30) and low numbers (near 23) of foci (Figure 1C). Strain mean MLH1 focus counts from females and males are uncorrelated (Spearman’s $\rho$ = `r sperman.corr.FM` -- 0.08; p = `r sperman.corr.FM.pval$p.value` -- 0.84) across the set of strains.

To further partition variation in recombination rate, we fit a series of linear models to mean MLH1 focus counts from `r length( DF.HetC.MixedModel.HQ$mean_co)` -- 137 house mice (from three subspecies; _M. m. domesticus_, _M. m. musculus_ and _M. m. molossinus_) (Table 2; detailed results available in Supplemental Table X). Strain, sex, subspecies, and sex\*subspecies each affect MLH1 focus count in a linear mixed model (M1; strain (random effect): p < `r 1/10000`-- 10-4; sex: p = `r MLH1.MMlmer_results$sex.results` -- 3.64 x 10-6; subspecies: p = `r MLH1.MMlmer_results$sub.sp` -- 9.69 x 10-4; subspecies\*sex: p = `r MLH1.MMlmer_results$interaction.pval` -- 1.8 x 10-4).

The effect of subspecies is no longer significant in a model treating all factors as fixed effects (M2; musculus p = `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][2]` -- 0.24, molossinus p = `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][3]` -- 0.1), highlighting strain and sex as salient variables. Two strains exhibit strong effects on MLH1 focus count (M3; *domesticus^G^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]` -- 1.78 x 10-6; *domesticus^LEW^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][4]` -- 0.02), with sex-strain interactions involving three strains (M3; *domesticus^G^* p < 10-6 -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][10]`; *molossinus^MSM^* p < 10-6 -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]`; *musculus^PWD^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]` -- 3.87 x 10-4). **fix the pvalue above, change the 0 to p < 10^-6**

In separate analyses of males (M4; n = `r length(male_mouse.avs_for.analysis$mouse)` -- 71), three strains disproportionately shape MLH1 focus count (as observed in Figure 1C): *musculus^PWD^* (p = `r sum_MLH1.male.glm.M4$coefficients[,4][7]` -- 3.6 x 10-7; effect = `r sum_MLH1.male.glm.M4$coefficients[7]` -- 6.11 foci **No CI's for these effect values**, *molossinus^MSM^* (p = `r sum_MLH1.male.glm.M4$coefficients[,4][8]` -- 6.3 x 10-9; effect = `r sum_MLH1.male.glm.M4$coefficients[8]`-- 6.91), and *musculus^SKIVE^* (p = -- 8.22 x 10-4; effect = -- 4.04). These three strains point to substantial evolution in the genome-wide recombination rate in spermatocytes; we subsequently refer to them as “high-recombination” strains. In females (M4; n= `r length(female_mouse.avs_for.analysis$mouse)` --76), three strains affect MLH1 focus count: *domesticus^G^* (p = `r sum_MLH1.female.spc.glm.M5$coefficients[,4][2]` -- 8.7 x 10-6; effect = `r sum_MLH1.female.spc.glm.M5$coefficients[2]` -- 3.3), *molossinus^MSM^* (p = `r sum_MLH1.female.spc.glm.M5$coefficients[,4][5]` -- 2.43 x 10-5; effect = `r sum_MLH1.female.spc.glm.M5$coefficients[5]` -- 2.99), and *domesticus^LEW^* (p = `r sum_MLH1.female.spc.glm.M5$coefficients[,4][3]`  -- 0.03; effect = `r sum_MLH1.female.spc.glm.M5$coefficients[3]` -- 1.69). Strain effect sizes in females are modest in magnitude compared to those in males. Together, these results demonstrate that the genome-wide recombination rate evolves in a highly sex-specific manner.

@@

The variation in sex differences in recombination we discovered provided an opportunity to determine whether sex differences in chromatin compaction, as measured by the length of the synaptonemal complex (SC), are reversed when heterochiasmy is reversed. In all strains except _musculus^SKIVE^_, females have longer SCs than males, whether SC length was estimated as the total length across bivalents or as the length of short bivalents (t-tests; all p < 0.05, except short bivalents in _musculus^SKIVE^_, p = `r Q1_short.SC_t.test.SKIVE$p.value` -- 0.11).
Among short bivalents (to which the female X bivalent does not contribute), female to male ratios of mouse mean SC length range from `r Q1SC.mus.av.short.biv_DIFF_PWD` -- 1.26 (*musculus^PWD^*) to `r Q1SC.mus.av.short.biv_DIFF_WSB` -- 1.52 (*domesticus^WSB^*) across strains. That females have longer SCs is further supported by models that include covariates, which identify sex as the most consistently significant effect for total SC length (M1: p = `r Q1.lmer_totSC_M1_results$sex.results` -- 2.56 x 10-31; M2: p = `r sum.Q1.glm_totSC_M2$coefficients[,4][8]` -- 2.56 x 10-8; M3: p =  `r sum.Q1.glm_totSC_M3$coefficients[,4][8]` -- 2.56 x 10-8) and short bivalent SC length (M1: p = `r Q1.lmer_short_M1_results$sex.results` -- 1.12 x 10-11; M2: p = `r summ.Q1.glm_short.SC_M2$coefficients[,4][8]/10000` -- 1.33 x 10-7; M3: p = `r summ.Q1.glm_short.SC_M3$coefficients[,4][8]/10000` -- 1.33 x 10-7). The existence of some subspecies and strain effects on total SC length and short bivalent SC length (Supplemental Tables 5 and 8) further indicates that SC length has evolved among strains and among subspecies. 
In summary, two approaches for measuring SC length demonstrate that females have longer SCs (chromosome axes), even in strains in which males recombine more. This pattern implies that in high-recombination strains, spermatocytes have less space than oocytes in which to position additional crossovers.

@@

We used normalized positions of MLH1 foci along bivalents with a single focus to compare crossover location while controlling for differences in SC length. In all strains, MLH1 foci tend to be closer to the telomere in males (mean normalized position in males: `r mean( (mouse.level.BivData_table %>% filter(sex == "male") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )` -- 0.68 **add s.d here?**; mean normalized position in females: `r mean( (mouse.level.BivData_table %>% filter(sex == "female") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )` --  0.56; paired t-test; p = `r Q1.strain.F1.Pos$p.value` -- 8.49 x 10-4). Sex is also the strongest determinant of MLH1 focus position in the models we tested (M1: p = `r lmer.Q1.nrm1CO_M1_results$sex.results` -- 2.82 x 10-26; M2: p = `r summry.Q1.glm_F1pos_M2$coefficients[,4][8]` -- 3.96 x 10-8; M3: p = `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][8]` -- 3.96 x 10-8).

Males have longer normalized mean inter-focal distances (IFD~norm~) than females in seven out of eight strains (t-tests; p < 0.05), with only _musculus^KAZ^_ showing no difference (p = `r Q1.IFD.PER_t.test_KAZ$p.value` -- 0.33). Examination of IFD~norm~ distributions indicates that females are centered at approximately 50% and show a slight enrichment of low (<25%) values, whereas males are enriched for higher values. Models treating IFD~norm~ as the dependent variable support the inference of stronger interference in males, with sex being the most significant variable (M1: p = `r Q1.IFD.PER_M1_lmer.results$sex.results` -- 9.08 x 10-12; M2: p = `r summry.glm.Q1_IFD.PER.M2$coefficients[,4][4]` -- 0.01; M3: p = `r summry.glm.Q1_IFD.PER.M3$coefficients[,4][2]` -- 0.01). In contrast, there is no clear signal of sex differences in raw mean inter-focal distances (IFD~raw~) (Supplemental Table 13 and 14) across the full set of strains, whether they are considered separately or together. Visualization of normalized MLH1 foci positions on bivalents with two crossovers **(Figure 4C and Supplemental Figure X)** further suggests that interference distances vary more in females than in males, and that males display a stronger telomeric bias in the placement of the distal crossover.

@@

Ninety-six percent of single bivalents in our pooled dataset (n = `r formatC( sum( Chrm.prop_strain_table$n0CO, Chrm.prop_strain_table$n1CO, Chrm.prop_strain_table$n2CO, Chrm.prop_strain_table$n3CO, Chrm.prop_strain_table$n4CO ), format="d", big.mark=",")`-- 9,569) have either one or two MLH foci (Supplemental Figure 2). The proportions of single-focus (1CO) bivalents vs. double-focus (2CO) bivalents distinguish high-recombination strains from low-recombination strains (Supplemental Figure 2). High-recombination strains are enriched for 2CO bivalents at the expense of 1CO bivalents: proportions of 2CO bivalents are `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "SKIVE"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "SKIVE"]` -- 0.33 in _musculus^SKIVE^_, `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "PWD"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "PWD"]` -- 0.44 in _musculus^PWD^_, and `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "MSM"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "MSM"]` -- 0.51 in _molossinus^MSM^_ (Supplemental Figure 3). Following patterns in the genome-wide recombination rate, male _musculus^PWD^_ and male _molossinus^MSM^_ have 2CO proportions that are more similar to each other than to strains from their own subspecies (chi-square tests; _musculus^PWD^_ vs. _molossinus^MSM^_: p = `r PWD.MSM.2CO.prop.test$p.value` -- 0.37; _musculus^PWD^_ vs. *musculus^KAZ^*: p = `r PWD.KAZ.2CO.prop.test$p.value` -- 1.23 x 10-31; _molossinus^MSM^_ vs. _molossinus^MOLF^_: p = `r MSM.MOLF.2CO.prop.test$p.value` -- 2.34 x  10-6). These results demonstrate that evolution of the genome-wide recombination rate reflects changes in crossover number across multiple bivalents.

To begin to localize evolution of genome-wide recombination rate to steps of the recombination pathway, we counted DMC1 foci in prophase spermatocytes as markers for double-strand breaks (DSBs). DMC1 foci were counted in a total of `r sum( (DMC1table.main$ncells.x) )` -- 76 early zygotene and `r sum( (DMC1table.main$ncells.y) )` -- 75 late zygotene spermatocytes from two high-recombination strains ( _musculus^PWD^_ and _molossinus^MSM^_) and three low-recombination strains (*musculus^KAZ^*, _domesticus^WSB^_, and _domesticus^G^_). High-recombination strains have significantly more DMC1 foci than low-recombination strains in early zygotene cells (t-test; -- `r 1/1000000`p < 10-6). In contrast, the two strain groups do not differ in DMC1 foci in late zygotene cells (t-test; p = `r ttest.HighLow.Z.pval` -- 0.66). Since DSBs are repaired as either COs or non-crossovers (NCOs), the ratio of MLH1 foci to DMC1 foci can be used to estimate the proportion of DSBs designated as COs. High-recombination and low-recombination strains do not differ in the MLH1/DMC1 ratio, whether DMC1 foci were counted in early zygotene cells or late zygotene cells (t-test; p > 0.05). These results raise the possibility that the evolution of genome-wide recombination rate is primarily determined by processes that precede the CO/NCO decision, at least in house mice.

Total SC length only partially differentiates high-recombination strains from low-recombination strains (Figure 3). Whereas high-recombination strains as a group have significantly greater total SC length than low-recombination strains (t-test; p = `r Q2tot.sc_t.test_High.Low$p.value` -- 0.01), separate tests within subspecies show that the two strain categories differ within *M. m. molossinus* (p = `r Q2tot.sc_t.test_High.Low_Mol$p.value` -- 2.59 x 10-4) but not within *M. m. musculus* (p = `r Q2tot.sc_t.test_High.Low_Musc$p.value` -- 0.65). Additionally, mouse means for the reduced (short and long) bivalent datasets do not differ between high-recombination and low-recombination strains (t-test; short: p = `r Q2short.biv_t.test_High.Low$p.value` -- 0.84; long: p = `r Q2long.biv_t.test_High.Low$p.value` -- 0.19). In a model with total SC length as the dependent variable (M4), the two subspecies effects are significant (*M. m. musculus* p = `r sum_totSC.male.glm.M4$coefficients[,4][2]` -- 3.95 x 10-7; *M. m. molossinus* p = `r sum_totSC.male.glm.M4$coefficients[,4][3]` -- 3.33 x 10-7), but there are also strain-specific effects (Supplemental Table 6). In models with SC lengths of short and long bivalents as dependent variables, several subspecies and strain effects reach significance (p < 0.05) (Supplemental Table 9), but they are not consistent across models. Collectively, these results reveal that evolution of SC length is not strongly associated with evolution of genome-wide recombination rate in house mice.

@@

**not many edits to replace, bbut double check**

We used a panel of wild-derived inbred strains of house mice (*Mus musculus*) and related murid species to profile natural genetic variation in recombination (Table 1). Our survey included 5 strains from *M. m. musculus*, 4 strains from *M. m. domesticus*, 2 strains from *M. m. molossinus*, 2 strains from *M. m. castaneus*, and 1 strain each from *M. spicilegus*, *M. spretus* and _M. caroli_. We subsequently denote strains by their abbreviated subspecies and name (e.g. *domesticus^WSB^*). Mice were housed at dedicated, temperature-controlled facilities in the UW-Madison School of Medicine and Public Health, with the exception of mice from Gough Island, which were housed in a temperature-controlled facility in the UW-Madison School of Veterinary Medicine. Mice from an inbred strain of Gough Island mice were sampled after 17 generations of brother-sister mating. All mice were provided with ad libitum food and water. Procedures followed protocols approved by IACUC. **statement on biological replicates**

@@

@@

Images were captured using a Zeiss Axioplan 2 microscope with AxioLab camera and AxioVision software (Zeiss, Cambridge, UK). Preprocessing, including cropping, noise reduction, and histogram adjustments, was performed using Photoshop (v13.0). Image file names were anonymized before manual scoring of MLH1 foci or DMC1 foci using Photoshop.

@@

To estimate the number of crossovers across the genome, we counted MLH1 foci. MLH1 foci were counted in cells with intact and complete karyotypes (19 acrocentric bivalents and XY for spermatocytes; 20 acrocentric bivalents for oocytes) and distinct MLH1 foci. A quality score ranging from 1 (best) to 5 (worst) was assigned to each cell based on visual appearance of staining and spread of bivalents. Cells with a score of 5 were excluded from the final analysis. Distributions of MLH1 count per cell were visually inspected for normality (Supplemental Figure 1). MLH1 foci located on the XY in spermatocytes were excluded from counts. In addition to MLH1 counts, we measured several traits to further characterize the recombination landscape. To estimate the number of double-strand breaks, a minority of which lead to crossovers, mean DMC1 foci per cell was quantified for a single male from each of a subset of strains (*molossinus^MSM^*, _musculus^PWD^_,  _domesticus^WSB^_, and _domesticus^G^_ ). SC morphology and CREST foci number were used to stage spermatocytes as early zygotene or late zygotene.

@@

This research was funded by NIH grants R01GM120051 and R01GM100426 to B. A. P.. A. L. P. was partly supported by NIH T32GM007133. Generous Franscico Peligri for microscope.

>>>>

Meiosis converts diploid germ cells into haploid gametes. During meiosis I, DNA crossovers aid the separation of homologous chromosomes by physically linking them and establishing tension between them on the spindle [@petronczki2003]. The wrong number of recombination events can disrupt chromosomal segregation, leading to infertility, miscarriage, and birth defects [@hassold2001]. Recombination also shapes evolution by shuffling the combinations of genetic variants offspring inherit. Recombination affects the fates of beneficial and deleterious mutations [@felsenstein1974; @fisher1930; @hill1966] and interacts with natural selection to leave gradients in genomic patterns of diversity [@begun1992; @nachman2012; @charlesworth1993; @cutter2013; @smith1974].

@@

We used counts of MLH1 foci per cell to estimate genome-wide recombination rates in `r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` wild-derived inbred strains sampled from three subspecies of house mice ( _M. m. domesticus_, _M. m. musculus_ and _M. m. molossinus_ ) and three other species of Mus ( _M. spretus_, _M. spicilegus_ , and _M. caroli_ ). Mean MLH1 focus counts for `r sum(main.table$Nmice)` mice were quantified from an average of `r mean( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T)` spermatocytes per male (for a total of `r formatC( sum( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T), format="d", big.mark=",")` spermatocytes) and `r mean( (Mouse.Table_HQ %>% filter(sex == "female") )$Ncells, na.rm = T)` oocytes per female (for a total of `r formatC(sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells), format="d", big.mark=",")` oocytes). **insert statement for cell sample size number**

Graphical comparisons reveal sex-specific dynamics to the evolution of genome-wide recombination rate (Figure 1A). First, MLH1 focus counts differ between females and males in most strains. Second, the difference in counts between the sexes varies among strains. Although most strains show more MLH1 foci in females, two strains (*musculus^PWD^* and _molossinus^MSM^_) exhibit higher counts in males. In females, numbers of MLH1 foci are evenly distributed around the sex-wide mean of approximately 25 (Figure 1B). In stark contrast, males largely separate into two groups of strains with high numbers (near 30) and low numbers (near 23) of foci (Figure 1C). Strain mean MLH1 focus counts from females and males are uncorrelated (Spearman’s $\rho$ = `r sperman.corr.FM`; p = `r sperman.corr.FM.pval$p.value`) across the set of strains.

To further partition variation in recombination rate, we fit a series of linear models to mean MLH1 focus counts from `r length( DF.HetC.MixedModel.HQ$mean_co)` house mice (from three subspecies; _M. m. domesticus_, _M. m. musculus_ and _M. m. molossinus_) (Table 2; detailed results available in Supplemental Table X). Strain, sex, subspecies, and sex\*subspecies each affect MLH1 focus count in a linear mixed model (M1; strain (random effect): p < `r 1/10000`-- 10-4; sex: p = `r MLH1.MMlmer_results$sex.results`; subspecies: p = `r MLH1.MMlmer_results$sub.sp`; subspecies\*sex: p = `r MLH1.MMlmer_results$interaction.pval`).

The effect of subspecies is no longer significant in a model treating all factors as fixed effects (M2; musculus p = `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][2]`, molossinus p = `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][3]`), highlighting strain and sex as salient variables. Two strains exhibit strong effects on MLH1 focus count (M3; *domesticus^G^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]`; *domesticus^LEW^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][4]`), with sex-strain interactions involving three strains (M3; *domesticus^G^* p < 10-6 -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][10]`; *molossinus^MSM^* p < 10-6 -- `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]`; *musculus^PWD^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]`). **fix the pvalue above, change the 0 to p < 10^-6**

In separate analyses of males (M4; n = `r length(male_mouse.avs_for.analysis$mouse)`), three strains disproportionately shape MLH1 focus count (as observed in Figure 1C): *musculus^PWD^* (p = `r sum_MLH1.male.glm.M4$coefficients[,4][7]`; effect = `r sum_MLH1.male.glm.M4$coefficients[7]` foci, *molossinus^MSM^* (p = `r sum_MLH1.male.glm.M4$coefficients[,4][8]`; effect = `r sum_MLH1.male.glm.M4$coefficients[8]`), and *musculus^SKIVE^* (p = `r sum_MLH1.male.glm.M4$coefficients[,4][9]`; effect = `r sum_MLH1.male.glm.M4$coefficients[9]`). These three strains point to substantial evolution in the genome-wide recombination rate in spermatocytes; we subsequently refer to them as “high-recombination” strains. In females (M4; n= `r length(female_mouse.avs_for.analysis$mouse)` ), three strains affect MLH1 focus count: *domesticus^G^* (p = `r sum_MLH1.female.spc.glm.M5$coefficients[,4][2]`; effect = `r sum_MLH1.female.spc.glm.M5$coefficients[2]`), *molossinus^MSM^* (p = `r sum_MLH1.female.spc.glm.M5$coefficients[,4][5]`; effect = `r sum_MLH1.female.spc.glm.M5$coefficients[5]`), and *domesticus^LEW^* (p = `r sum_MLH1.female.spc.glm.M5$coefficients[,4][3]`; effect = `r sum_MLH1.female.spc.glm.M5$coefficients[3]`). Strain effect sizes in females are modest in magnitude compared to those in males. Together, these results demonstrate that the genome-wide recombination rate evolves in a highly sex-specific manner.

@@

The variation in sex differences in recombination we discovered provided an opportunity to determine whether sex differences in chromatin compaction, as measured by the length of the synaptonemal complex (SC), are reversed when heterochiasmy is reversed. In all strains except _musculus^SKIVE^_, females have longer SCs than males, whether SC length was estimated as the total length across bivalents or as the length of short bivalents (t-tests; all p < 0.05, except short bivalents in _musculus^SKIVE^_, p = `r Q1_short.SC_t.test.SKIVE$p.value`).
Among short bivalents (to which the female X bivalent does not contribute), female to male ratios of mouse mean SC length range from `r Q1SC.mus.av.short.biv_DIFF_PWD` (*musculus^PWD^*) to `r Q1SC.mus.av.short.biv_DIFF_WSB` (*domesticus^WSB^*) across strains. That females have longer SCs is further supported by models that include covariates, which identify sex as the most consistently significant effect for total SC length (M1: p = `r Q1.lmer_totSC_M1_results$sex.results`; M2: p = `r sum.Q1.glm_totSC_M2$coefficients[,4][8]`; M3: p =  `r sum.Q1.glm_totSC_M3$coefficients[,4][8]`) and short bivalent SC length (M1: p = `r Q1.lmer_short_M1_results$sex.results`; M2: p = `r summ.Q1.glm_short.SC_M2$coefficients[,4][8]`; M3: p = `r summ.Q1.glm_short.SC_M3$coefficients[,4][8]`). The existence of some subspecies and strain effects on total SC length and short bivalent SC length (Supplemental Tables 5 and 8) further indicates that SC length has evolved among strains and among subspecies.

In summary, two approaches for measuring SC length demonstrate that females have longer SCs (chromosome axes), even in strains in which males recombine more. This pattern implies that in high-recombination strains, spermatocytes have less space than oocytes in which to position additional crossovers.

@@

We used normalized positions of MLH1 foci along bivalents with a single focus to compare crossover location while controlling for differences in SC length. In all strains, MLH1 foci tend to be closer to the telomere in males (mean normalized position in males: `r mean( (mouse.level.BivData_table %>% filter(sex == "male") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )`; mean normalized position in females: `r mean( (mouse.level.BivData_table %>% filter(sex == "female") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )`; paired t-test; p = `r Q1.strain.F1.Pos$p.value`). Sex is also the strongest determinant of MLH1 focus position in the models we tested (M1: p = `r lmer.Q1.nrm1CO_M1_results$sex.results`; M2: p = `r summry.Q1.glm_F1pos_M2$coefficients[,4][8]`; M3: p = `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][8]`).

Males have longer normalized mean inter-focal distances (IFD~norm~) than females in seven out of eight strains (t-tests; p < 0.05), with only _musculus^KAZ^_ showing no difference (p = `r Q1.IFD.PER_t.test_KAZ$p.value`). Examination of IFD~norm~ distributions indicates that females are centered at approximately 50% and show a slight enrichment of low (<25%) values, whereas males are enriched for higher values. Models treating IFD~norm~ as the dependent variable support the inference of stronger interference in males, with sex being the most significant variable (M1: p = `r Q1.IFD.PER_M1_lmer.results$sex.results`; M2: p = `r summry.glm.Q1_IFD.PER.M2$coefficients[,4][4]`; M3: p = `r summry.glm.Q1_IFD.PER.M3$coefficients[,4][2]`). In contrast, there is no clear signal of sex differences in raw mean inter-focal distances (IFD~raw~) (Supplemental Table 13 and 14) across the full set of strains, whether they are considered separately or together. Visualization of normalized MLH1 foci positions on bivalents with two crossovers **(Figure 4C and Supplemental Figure X)** further suggests that interference distances vary more in females than in males, and that males display a stronger telomeric bias in the placement of the distal crossover.

@@

Ninety-six percent of single bivalents in our pooled dataset (n = `r formatC( sum( Chrm.prop_strain_table$n0CO, Chrm.prop_strain_table$n1CO, Chrm.prop_strain_table$n2CO, Chrm.prop_strain_table$n3CO, Chrm.prop_strain_table$n4CO ), format="d", big.mark=",")`) have either one or two MLH foci (Supplemental Figure 2). The proportions of single-focus (1CO) bivalents vs. double-focus (2CO) bivalents distinguish high-recombination strains from low-recombination strains (Supplemental Figure 2). High-recombination strains are enriched for 2CO bivalents at the expense of 1CO bivalents: proportions of 2CO bivalents are `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "SKIVE"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "SKIVE"]` in _musculus^SKIVE^_, `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "PWD"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "PWD"]` in _musculus^PWD^_, and `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "MSM"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "MSM"]` in _molossinus^MSM^_ (Supplemental Figure 3). Following patterns in the genome-wide recombination rate, male _musculus^PWD^_ and male _molossinus^MSM^_ have 2CO proportions that are more similar to each other than to strains from their own subspecies (chi-square tests; _musculus^PWD^_ vs. _molossinus^MSM^_: p = `r PWD.MSM.2CO.prop.test$p.value`; _musculus^PWD^_ vs. *musculus^KAZ^*: p = `r PWD.KAZ.2CO.prop.test$p.value`; _molossinus^MSM^_ vs. _molossinus^MOLF^_: p = `r MSM.MOLF.2CO.prop.test$p.value`). These results demonstrate that evolution of the genome-wide recombination rate reflects changes in crossover number across multiple bivalents.

To begin to localize evolution of genome-wide recombination rate to steps of the recombination pathway, we counted DMC1 foci in prophase spermatocytes as markers for double-strand breaks (DSBs). DMC1 foci were counted in a total of `r sum( (DMC1table.main$ncells.x) )` early zygotene and `r sum( (DMC1table.main$ncells.y) )` late zygotene spermatocytes from two high-recombination strains ( _musculus^PWD^_ and _molossinus^MSM^_) and three low-recombination strains (*musculus^KAZ^*, _domesticus^WSB^_, and _domesticus^G^_). High-recombination strains have significantly more DMC1 foci than low-recombination strains in early zygotene cells (t-test; p < `r 1/1000000` -- 10-6). In contrast, the two strain groups do not differ in DMC1 foci in late zygotene cells (t-test; p = `r ttest.HighLow.Z.pval`). Since DSBs are repaired as either COs or non-crossovers (NCOs), the ratio of MLH1 foci to DMC1 foci can be used to estimate the proportion of DSBs designated as COs. High-recombination and low-recombination strains do not differ in the MLH1/DMC1 ratio, whether DMC1 foci were counted in early zygotene cells or late zygotene cells (t-test; p > 0.05). These results raise the possibility that the evolution of genome-wide recombination rate is primarily determined by processes that precede the CO/NCO decision, at least in house mice.

Total SC length only partially differentiates high-recombination strains from low-recombination strains (Figure 3). Whereas high-recombination strains as a group have significantly greater total SC length than low-recombination strains (t-test; p = `r Q2tot.sc_t.test_High.Low$p.value`), separate tests within subspecies show that the two strain categories differ within *M. m. molossinus* (p = `r Q2tot.sc_t.test_High.Low_Mol$p.value`) but not within *M. m. musculus* (p = `r Q2tot.sc_t.test_High.Low_Musc$p.value`). Additionally, mouse means for the reduced (short and long) bivalent datasets do not differ between high-recombination and low-recombination strains (t-test; short: p = `r Q2short.biv_t.test_High.Low$p.value`; long: p = `r Q2long.biv_t.test_High.Low$p.value`). In a model with total SC length as the dependent variable (M4), the two subspecies effects are significant (*M. m. musculus* p = `r sum_totSC.male.glm.M4$coefficients[,4][2]`; *M. m. molossinus* p = `r sum_totSC.male.glm.M4$coefficients[,4][3]`), but there are also strain-specific effects (Supplemental Table 6). In models with SC lengths of short and long bivalents as dependent variables, several subspecies and strain effects reach significance (p < 0.05) (Supplemental Table 9), but they are not consistent across models. Collectively, these results reveal that evolution of SC length is not strongly associated with evolution of genome-wide recombination rate in house mice.

@@

**not many edits to replace, but double check**

We used a panel of wild-derived inbred strains of house mice (*Mus musculus*) and related murid species to profile natural genetic variation in recombination (Table 1). Our survey included 5 strains from *M. m. musculus*, 4 strains from *M. m. domesticus*, 2 strains from *M. m. molossinus*, 2 strains from *M. m. castaneus*, and 1 strain each from *M. spicilegus*, *M. spretus* and _M. caroli_. We subsequently denote strains by their abbreviated subspecies and name (e.g. *domesticus^WSB^*). **Each mouse from a inbred group was treated as a biological replicate**.

Mice were housed at dedicated, temperature-controlled facilities in the UW-Madison School of Medicine and Public Health, with the exception of mice from Gough Island, which were housed in a temperature-controlled facility in the UW-Madison School of Veterinary Medicine. Mice were sampled from an partially-inbred strain of Gough Island mice, approximately 6 generations of brother-sister matting. All mice were provided with ad libitum food and water. Procedures followed protocols approved by IACUC.

@@

**technical replicate definition** 
Three slides of cell spreads per mouse were prepared to serve as technical replicates for the staining protocol. Comparisons of multiple stained slides from the same mouse, showed no difference in mean MLH1 cell counts and mean cell quality.

@@

The number of cells imaged per individual mouse is based on previous studies [Dumont, Wang]. Images were captured using a Zeiss Axioplan 2 microscope with AxioLab camera and AxioVision software (Zeiss, Cambridge, UK). Preprocessing, including cropping, noise reduction, and histogram adjustments, was performed using Photoshop (v13.0). Image file names were anonymized before manual scoring of MLH1 foci or DMC1 foci using Photoshop.

@@

To estimate the number of crossovers across the genome, we counted MLH1 foci. MLH1 foci were counted in cells with intact and complete karyotypes (19 acrocentric bivalents and XY for spermatocytes; 20 acrocentric bivalents for oocytes) and distinct MLH1 foci. A quality score ranging from 1 (best) to 5 (worst) was assigned to each cell based on visual appearance of staining and spread of bivalents. Cells with a score of 5 were excluded from the final analysis. Distributions of MLH1 count per cell were visually inspected for normality (Supplemental Figure 1). **During preliminary analysis, cells found to be outliers for MLH1 count had the original image was examined for issues.**

MLH1 foci located on the XY in spermatocytes were excluded from counts. In addition to MLH1 counts, we measured several traits to further characterize the recombination landscape. To estimate the number of double-strand breaks, a minority of which lead to crossovers, mean DMC1 foci per cell was quantified for a single male from each of a subset of strains (*molossinus^MSM^*, _musculus^PWD^_,  _domesticus^WSB^_, and _domesticus^G^_ ). SC morphology and CREST foci number were used to stage spermatocytes as early zygotene or late zygotene.

@@

This research was funded by NIH grants R01GM120051 and R01GM100426 to B. A. P.. A. L. P. was partly supported by NIH T32GM007133. We are grateful to Francisco Pelegri for generous assistance with microscopy.
